Template for indicator documentation

### 2023-08-02 functional_plant_indicators_seminat.Rmd
Update functional_plant_indicators_seminat.Rmd
corrected some minor typos in the first few paragraphs

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for semi-natural ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the moisture signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

@@

- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019, contains semi-natural grasslands(T32), semi-natural wet meadows (V10), heathlands (T34), and we also decided to include near-semi-natural meadows (T41) ) and (2) from the ASO monitoring scheme (cf. Bär et al. 2021, targets semi-natural grasslands (T32) only)
- as reference data we use (3) generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- (4) Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)

@@

(2) ASO monitoring data
ASO stands for 'arealrepresentativ overvåking av semi-naturlig eng', i.e. 'area representative monitoring of semi-natural meadows'. In ASO ca. 20 areas are chosen annually based on a probabilistic distribution model of semi-natural meadows in Norway (see Johansen et al. 2017 for details). Each ASO area (område) spans a 10 x 10 km grid cell, in which a maximum of up to 10 sites (flate) of 500 x 500 m are chosen. In each site, all semi-natural meadows (eng) with an area of at least 250 sqm are delineated. In up to three meadows per site the plant species composition is sampled using a transect method. For the vegetation transect analyses, both cover and frequency of each vascular plant species along the transect is recorded. For every meadow, the species data collection is accompanied by an assessment of the ecosystem, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
In the analysis in this document, we use the ASO data from 2022 only as the data from 2021 and the ASO pilot in 2019 are not in sufficiently good shape. In the future, all ASO data will hopefully be available in Miljødirektoratets kartkatalog.

@@

(4) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance for semi-natural ecosystems as these are thought to be subject to potential change due to abandonment, drainage/flooding, pollution, and erosion.

>>>>

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for semi-natural ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH, nitrogen, phosphorus, grazing/mowing, and soil disturbance. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the moisture signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

@@

- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019, contains semi-natural grasslands T32, semi-natural wet meadows V10, heathlands T34, and we also decided to include near-semi-natural meadows T41) and (2) from the ASO monitoring scheme (cf. Bär et al. 2021, targets semi-natural grasslands T32 only)
- as reference data we use (3) generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- finally, (4) Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)

@@

(2) ASO monitoring data:
ASO stands for 'arealrepresentativ overvåking av semi-naturlig eng', i.e. 'area representative monitoring of semi-natural meadows'. In ASO ca. 20 areas are chosen annually based on a probabilistic distribution model of semi-natural meadows in Norway (see Johansen et al. 2017 for details). Each ASO area (område) spans a 10 x 10 km grid cell, in which a maximum of up to 10 sites (flate) of 500 x 500 m are chosen. In each site, all semi-natural meadows (eng) with an area of at least 250 sqm are delineated. In up to three meadows per site the plant species composition is sampled using a transect method. For the vegetation transect analyses, both cover and frequency of each vascular plant species along the transect is recorded. For every meadow, the species data collection is accompanied by an assessment of the ecosystem, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
In the analysis in this document, we use the ASO data from 2022 only as the data from 2021 and the ASO pilot in 2019 are not in sufficiently good shape. In the future, all ASO data will hopefully be available in Miljødirektoratets kartkatalog.

@@

(4) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, nitrogen, phosphorus, grazing/mowing, and soil disturbance for semi-natural ecosystems, as these are thought to be subject to potential change due to abandonment, drainage/flooding, pollution, and erosion.
